---
title: "TDA and TDAmapper"
author: "Kaisa Taipale"
date: "November 30, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is TDA?

Topological Data Analysis (TDA) is a way of looking at the topology of data, rather than linear fits or probability models. What's the **shape** of data?

Pros: shape persists however you "rotate" your data (point of view matters less).


## Today: TDA with R

- Example of a circle with TDA and TDAmapper packages
- Example of an infinity symbol
- Stock data (messy, real life)

## Packages
```{r packagestouse, echo=TRUE, message=FALSE}
library(BatchGetSymbols)
library(reshape2)
library(TDA)
library(TDAmapper)
library(igraph)
```
## Circle first
```{r circle, echo = TRUE}
circle1 <- circleUnif(60)
circle2 <- circleUnif(60,r=2)+3
circles <- rbind(circle1,circle2)
plot(circles)
```


## Circle persistence diagram
Persistence diagram: birth and death of homology classes
```{r circlepersistent}
circleDiag<-ripsDiag(circles,maxscale=4, maxdimension = 1,library = "GUDHI")
plot(circleDiag[["diagram"]])
```

## TDAmapper for the circle

```{r ciclemapperex, echo=TRUE, message=FALSE}
circlemapper <- mapper1D(distance_matrix=dist(circles), filter_values = circles[,1], 
  num_intervals =  12, 
  percent_overlap = 50,
  num_bins_when_clustering =12)
g2 <- graph.adjacency(circlemapper$adjacency,   
  mode="undirected")
```
## Plot result from TDAmapper for circles
```{r resultforcircles, echo=TRUE, message=FALSE}
plot(g2, layout = layout.auto(g2) )
```

## Infinity symbol
```{r infinity, echo = TRUE}
infinitydf<- cbind(2*cos(0.5*(1:100)),sin(1:100))
plot(infinitydf)
```

## Persistence diagram
```{r infintypersistence}
infinityDiag<-ripsDiag(infinitydf,
  maxscale = 1.6, 
  maxdimension = 1)
plot(infinityDiag[["diagram"]])
```

## TDAmapper

```{r infinitymapperexample, echo=TRUE}
infinitymapper <- mapper1D(distance_matrix=dist(infinitydf), 
  filter_values = infinitydf[,1], 
  num_intervals =  12, 
  percent_overlap = 50,
  num_bins_when_clustering =12)

g2 <- graph.adjacency(infinitymapper$adjacency, 
  mode="undirected")
```


## TDAmapper for infinity symbol
```{r plotfortheinfinity, echo=FALSE}
plot(g2, layout = layout.auto(g2) )
```

## Financial data!
- With students Jacqueline Cai and Hao Li at the University of Minnesota
- Inspired by paper "Topology Data Analysis Of Critical Transitions in
Financial Networks" by Marian Gidea
- Visualize transition in stock correlations before crisis (?)


## Get stock data and make log returns matrix

```{r stockdata, echo = TRUE, message=FALSE}
first.date <- as.Date('2004/01/01')
last.date <- as.Date('2009/09/30')
tickers <- c('MMM','AXP','AAPL','BA','CAT','CVX','CSCO','KO','DWDP','XOM','GE','GS','HD','IBM','INTC','JNJ','JPM','MCD','MRK','MSFT','NKE','PFE','PG','TRV','UNH','UTX','VZ','WMT','DIS')
data <- BatchGetSymbols(tickers = tickers,
                        first.date = first.date,
                        last.date = last.date)
newdata <- BatchGetSymbols(tickers = tickers,
            first.date = as.Date('2017/05/01'),
            last.date = as.Date('2017/10/01'))
widestock <- dcast(data$df.tickers[,6:8], ref.date ~ ticker, value.var="price.adjusted")
logrets <- apply(widestock[,2:30], 2, 
                function(x) diff(log(x), lag=1))
```

## Make a correlation matrix for the log returns
Correlation matrices are really easy!
```{r correlationmatrix, echo=TRUE}
first100cor <- cor(logrets[1:100,], method = "pearson")
second100cor <- cor(logrets[101:200,], method = "pearson")
```
## Graphing the correlation networks

Let's make a matrix by looking at a "super-level set," basically only showing edges when stocks have highly correlated log returns. Notice I'm making a "distance" matrix out of the correlation matrix.
```{r superlevel100, echo=TRUE}
highcor100mat <- sqrt(2*(first100cor+1))
highcorsuperlevel100 <- ifelse (highcor100mat>1.8, highcor100mat,0)
network=graph_from_adjacency_matrix(highcorsuperlevel100, 
        weighted=TRUE, 
        mode="undirected", 
        diag=F)

```
## Plot for first 100 days
```{r plotfirst100, echo=TRUE}
plot(network)
```


## Plot for second 100 days
```{r superlevel200}
highcor200mat <- sqrt(2*(second100cor+1))
highcorsuperlevel200 <- ifelse (highcor200mat>1.8, highcor200mat,0)
network=graph_from_adjacency_matrix(highcorsuperlevel200, 
          weighted=TRUE, 
          mode="undirected", 
          diag=F)
plot(network)
```


## Financial crisis 100 days
Look now at 100 days starting December 21, 2007.
```{r superlevel1000}
later1000cor <- cor(logrets[1001:1100,], method = "pearson")
highcor1000mat <- sqrt(2*(later1000cor+1))
highcorsuperlevel1000 <- ifelse (highcor1000mat>1.8, highcor1000mat,0)
network=graph_from_adjacency_matrix(highcorsuperlevel1000, weighted=TRUE, mode="undirected", diag=F)
plot(network)
```

## Financial crisis 1200 days
And now October-November-December 2008
```{r superlevel1200}
later1200cor <- cor(logrets[1201:1300,], method = "pearson")
highcor1200mat <- sqrt(2*(later1200cor+1))
highcorsuperlevel1200 <- ifelse (highcor1200mat>1.8, highcor1200mat,0)
network=graph_from_adjacency_matrix(highcorsuperlevel1200, weighted=TRUE, mode="undirected", diag=F)
plot(network)
```

## Persistence diagram for stock data
We can make a persistence diagram for the stock data as well, to see if any "circles" exist or persist.
```{r stockdatapersistent, echo=TRUE}
first100persistence <- ripsDiag(highcor100mat,maxscale=2,maxdimension = 1,dist="arbitrary")

plot(first100persistence[["diagram"]])

```


## The next 200 days
Persistence diagram for the next 200 days, still in 2004-2005.
```{r nextpersistencediag}
second100persistence <- ripsDiag(highcor200mat,maxscale=2,maxdimension = 1,dist="arbitrary")
plot(second100persistence[["diagram"]])
```

## 1001-1100 days
Persistence diagram for the network from 2007.
```{r 2007persistencediag}
crisispersistence <- ripsDiag(highcor1000mat,maxscale=2,maxdimension = 1,dist="arbitrary")
plot(crisispersistence[["diagram"]])
```

## 1001-1100 days
Persistence diagram for the network from 2007.
```{r 2008persistencediag}
crisis2persistence <- ripsDiag(highcor1200mat,maxscale=2,maxdimension = 1,dist="arbitrary")
plot(crisis2persistence[["diagram"]])
```

## What about now? 

```{r now, echo=FALSE}

newwidestock <- dcast(newdata$df.tickers[,6:8],ref.date~ticker, value.var="price.adjusted")
newlogrets <- apply(widestock[,2:30],2, function(x) diff(log(x), lag=1))
newcor<-cor(newlogrets, method="pearson")

highcornew <-sqrt(2*(newcor+1))
highcorsuperlevel <-ifelse(highcornew>1.8,highcornew, 0)
network = graph_from_adjacency_matrix(highcorsuperlevel, weighted=TRUE, mode="undirected", diag=F)
plot(network)
```


